Iterative Prompting: Programmatic Data Exploration for Non-programmers

### 2021-04-06 wip-design/paper.tex
Finish cleanup for resubmission

\begin{document}
\title{The Gamma: Data Exploration through Iterative Prompting}

@@

\begin{abstract}
  Data exploration tools based on code have many desirable characteristics. They can easily access
  a wide range of different data sources, result in reproducible scripts and encourage users to
  reuse and modify existing data analyses. Unfortunately, most programming tools require expert
  coding skills. Can we make data exploration based on code accessible to non-experts?
  %
  We present The Gamma, a novel text-based data exploration environment.
  The Gamma is based on a single interaction principle and using it results in transparent and
  reproducible scripts. This lowers the barrier to entry and allows learning from previously
  created data analyses. We evaluate the usability and learnability of The Gamma through a user
  study on non-technical employees of a research institute.
  %
  Our work shows that we may not need to shy away from code in order to build accessible,
  reproducible and transparent tools that will allow a broad audience to benefit from the
  rise of open data.
\end{abstract}

@@

\begin{figure}
\includegraphics[width=.58\columnwidth]{figures/thegamma-annot}
\vspace{-0.5em}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechanism
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\vspace{-0.5em}
\end{figure}

@@

\section{Introduction}
Data science has more capabilities to help us understand the world than ever before, yet at the
same time post-truth politics and increasing public distrust in statistics makes data-driven insights
increasingly less relevant in public discourse~\cite{howstatslost}. To reverse this trend, we
need tools that let non-experts, including journalists and other information-literate citizens,
produce transparent, engaging data analyses that are easy to interpret without requiring expert
programming skill~\cite{ddj}. The design of such data exploration tool poses a unique mix of challenges.
First, the tool needs to have a very low barrier to entry. Second, it needs to support a wide
range of data sources in a uniform way. Third, the resulting data analyses need to assist
readers in learning how to reproduce the work and verify the claims it makes.

@@

\subsubsection*{Iterative Prompting}
The main contribution of our work is the \emph{iterative prompting} interaction principle,
which makes it possible to construct all valid data exploration scripts by repeatedly choosing
from a list of offered options. This way, non-programmers can write entire scripts through
auto-complete without learning a programming language first, but still produce transparent
and reproducible code. In other words, iterative prompting turns auto-complete from a programmer
assistance tool into a non-expert programming mechanism. A crucial feature is that iterative
prompting only offer operations that are valid in a given context and that it offer all such
operations; it is both correct and complete.

\subsubsection*{Data Exploration}
The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
The user accesses data available in a structured format. They make several experiments to find an
interesting way of looking at the data, e.g.~by applying different aggregations or filters. They
may choose to view the results as a table or a basic chart before publishing their analysis. The
Gamma makes such programmatic data exploration simple enough for non-experts, but scraping and
cleaning of messy data or building custom data visualizations is currently outside of the scope of
our work. Exposing those using iterative prompting remains an interesting future challenge.

\subsubsection*{Paper Overview}
The Gamma is available (non-anonymously) at \url{http://thegamma.net},
both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
evaluate the design principles behind the project\footnote{This paper is X pages, without citations,
to be commensurate with the size of the contribution.}:

@@

\item We reflect how our design lowers barriers to entry (Section~\ref{sec:design-bar}),
  supports learning without experts (Section~\ref{sec:design-expert}) and offers a
  complete and correct program construction method (Section~\ref{sec:design-cc}).

\item We evaluate the system through a number of case studies (Section~\ref{sec:cases})
  and a user study (Section~\ref{sec:study}), which confirms  that non-programmers can use The Gamma
  to construct non-trivial data queries.
\end{itemize}

@@

\noindent
We start from the \ikvd{Doctor} node and then follow two relationships. We use
\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
to follow links to all enemies of the Doctor and then specify
\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
to select only enemies that appear in a specific episode. The resulting query
is illustrated in Figure~\ref{fig:graphtp}.

The provider works with any graph database and generates members automatically, based on the
data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
relationships. The results returned by the provider is a table of properties of all nodes
along the specified path. As illustrated by an example discussed in Section~\ref{sec:cases},
the returned table can be further queried using the tabular data type provider.

\begin{figure}[t]
\centering
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-dr}
  \caption{Exploring Dr Who graph database by composing type providers}
  \label{fig:cases-dr}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-mp}
  \caption{Exploring Olympic medallists using tabular data type provider}
  \label{fig:cases-mp}
\end{subfigure}
\vspace{-0.5em}
\caption{Charts produced by two case studies of using The Gamma.}
\label{fig:cases}
\end{figure}

@@

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data explorations. We assess the simplicity of The Gamma
in Section~\ref{sec:study}. Here, we consider two case studies that evaluate expressivity and show what
can be achieved using the simple iterative prompting principle\footnote{Available (non-anonymously) at:
\url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
We used The Gamma for larger projects that analyse the UK government expenditure, present
activities of a research institute, analyse Olympic medals and explore information about the Doctor
Who series\footnote{The analyses are available (non-anonymously)
at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.

@@

Our objective is to develop a text-based data exploration tool that can be used by non-programmers.
Using the characterization by Olsen~\cite{evaluating}, data exploration environments are complex
systems that do not yield to simple controlled experimentation. We discuss the limits of evaluating
The Gamma in Section~\ref{sec:discussion-eval} and do not attempt to make make quantitative comparison
e.g.~with SQL or Tabelau~\cite{tableau}. Consequently, our goals are more modest.

\subsection{Study Design}
We evaluate the extent to which non-programmers can use The Gamma to conduct
text-based data exploration (\emph{RQ1}). To test this, we gave volunteers one of four data
exploration tasks and assessed whether they were able to complete the task, as well as how much
assistance, if any, they needed.
%
Some aspects of the study aim to shed light on three questions concerning learnability
of The Gamma, in particular whether knowledge can be trasferred between different data sources
(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
language used in the tabular data source (\emph{RQ2c}).

We performed a between-subjects study to evaluate the first experience of using The Gamma.
We recruited 13 participants (5 male, 8 female) from a business team
of a research lab working in non-technical roles (project management,
partnerships, communications) including one former journalist.  Only one participant (\#12)
had prior programming experience. We split participants into 4 groups. We gave participants a
brief overview of The Gamma (with content depending on the task) and then asked participants
to complete a task. The participants worked on the tasks for 30 minutes, after which we conducted
a 30 minute semi-structured group interview. We let participants work independently, but offered
guidance if they got stuck and asked for help. The four tasks were:
% The participants were volunteers who responded to an invitation sent to an internal mailing list.

\begin{itemize}
\item \emph{Expenditure.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{expenditure} data source to compare the UK government spending
  on ``Public order and safety'' and ``Defence'' in terms of GDP.
\item \emph{Lords.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{lords} data source (a table with House of Lords
  members expenses) to find a members representing London with the highest travel costs.
\item \emph{Worldbank.} Participants were given a minimal demo of iterative prompting and
  a code sample using \emph{worldbank}. They were asked to solve a different task using
  the \emph{worldbank} data source.
\item \emph{Olympics.} Participants were given a demo using \emph{olympics}.
  They were asked to solve a more complex problem, involving grouping and aggregation,
  using the same data source.
\end{itemize}

@@

\noindent
All of the tasks aim to answer the primary question RQ1 for one of the supported data sources.
In addition, the tasks \emph{expenditure} and \emph{lords} study RQ2a because they use different
data source in introduction and taks; \emph{worldbank} study RQ2b in that it comes with only brief
introduction and \emph{olumpics} studies RQ2c using our most complex data source.

\begin{table}
\centering
\begin{tabular}{l l l c l}
  \toprule
    & {\small \textit{Task}}
    & {\small \textit{Kind}} & {\small \textit{Done}}
    & {\small \textit{Notes}} \\
  \midrule
  \small \#1\qquad\qquad & \small expenditure \qquad\qquad & \small cube \qquad\qquad & \qquad\priority{50}\qquad\qquad & {\small Obtained one of two data series}\\
  \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
  \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
  \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
  \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
  \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
  \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
  \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
  \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
  \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
  \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
  \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
  \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
  \bottomrule
\end{tabular}

@@

\subsection{Study Results}
Table~\ref{tab:tasks} summarizes the work done by the study participants. Most notably, all
participants were able to complete, at least partially, a non-trivial data exploration task and
only half of them required further guidance. For each participant, we record the task, the kind
of data source used in the task and the level of completion. For participants who needed assistance,
the notes section details the help given. The experience suggests a number of possible design
improvements for The Gamma, which are discussed below.

\subsubsection*{Can Non-programmers Explore Data with The Gamma?}
Three facts allow us to answer RQ1 in the affirmative. First, 9 out of 11 participants agree or
strongly agree that they ``found the system easy to use''. Second, participants spent 10--25 minutes
(average 17 minutes) working with The Gamma and 12 out of 13 completed the task; 6 required assistance,
but 3 of those faced one issue (discussed later) that could be addressed in the introduction.
Third, a number of participants shared positive comments in the interviews.
For example, participant \#3 noted taht \emph{``this is actually pretty simple to use.
You think about the logic of what you're actually asking and then you try to get it into the format you can.''}
Participant \#2 noted that The Gamma alleviated their unease about code:
\emph{``For somebody who does not do coding or programming, this does not feel that daunting.
  It's not like you're giving me large screen full of code, which is reassuring.''}

@@

\subsubsection*{How Users Learn The Gamma?}
The nature of RQ2 makes it a challenging question to study and finding conclusive answers
may require further research. There is some evidence that knowledge can be transferred between
different data sources (RQ2a). Two of the tasks (\emph{expenditure} and \emph{lords}) used
one data source in the introduction and another for the task. Participants were able to complete
those, although \emph{lords} has been more challenging as it involves a complex data source.
Participant \#2 also gives a positive answer \emph{``I found it quite easy to translate what you
showed us in the demo to the new dataset.''}.

There is also some evidence that users can learn just from code samples (RQ2b). In the
\emph{worldbank} task, participants were given only a minimal demo of how to invoke iterative
prompting together with print-out of 2 code samples. All three participants were able to complete
a related task using the same data source. When discussing suitable educational materials for
The Gamma, participant \#7 also confirmed that having code is sufficient when they noted that
\emph{``a video would just be this [i.e.~a code sample] anyway''}. This supports our hypothesis
that, once a user understands the iterative prompting principle, they can learn how to use any
specific data source just from code samples.

\subsubsection*{How Users Understand Complex Query Languages?}
Although iterative prompting itself is simple, specific data sources may
expose richer structure. To ascertain their understanding of the tabular type provider (RQ2c), we
asked participants who worked with tabular data (\emph{lords} or \emph{olympics}) about their
understanding of the  \ikvd{then} member. This is a regular member, but it has a
special meaning in the domain specific language. For example, when specifying a list of
aggregations, the user can repeatedly select further aggregations. Selecting \ikvd{then} completes
the operation and allows the user to choose other operations such as sorting.

Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
over multiple lines, but rejected the idea after experimenting.
Participant \#12 then correctly concluded that it \emph{``allows us
to chain together the operations''} of the query; after a hint, participant \#13 reflected
that \emph{``if I knew this from the start, it would [have been easier].''} This summarizes an
important fact. Iterative prompting allows the users to start exploring new data sources, but
the structures exposed by more complex data sources have their own further design principles that
the users need to understand.

\subsubsection*{Making Complex Things Possible May Hurt}
The Gamma supports operations such as \ikvd{take(5)}. Most type
providers never generate those, but the provider for working with tabular data is an exception.
When filtering data, the provider allows specifying a condition on numerical attributes such as
\ikvd{olympics.\textquotesingle filter data\textquotesingle.\textquotesingle Year is greater than\textquotesingle(2004)}.
%
Three participants (\#11, \#12, \#13) struggled to complete a task, because they
initially attempted to use those operations. However, those operations violate the iterative prompting
principle as one cannot type `.' after \ikvd{\textquotesingle Year is greater than\textquotesingle}.
This suggests that we should either avoid such operations, or hide them under an ``advanced operations''
tab as a caution to novice users.

\subsubsection*{Benefits and Drawbacks of Text}
The Gamma is based on text to aid transparency. Text implies that there is no hidden state and
the reader sees the full code. The study suggests that using a text editor has both
benefits and drawbacks compared to alternatives such as structured editors~\cite{structure-based,livenut,lamdu}.
Most participants had no difficulty navigating around source code, making edits or deleting code
fragments, which is arguably harder in an unfamiliar structured editor.

On the one hand, we observed two issues in the study. Participant \#2 struggled with correct indentation, starting
a second command with more indentation than needed and participant \#6 had a syntax error in an unrelated command,
which prevents charts from rendering. On the other hand, some participants used the text editor effectively,
e.g.~participant \#5, who used copy-and-paste to fetch the same data series for multiple countries.

@@

\section{Design Reflections}
\label{sec:design}

The design of The Gamma has been motivated by a curiosity as to whether iterative
prompting can make text-based programming with data accessible to non-experts. In this section,
we theoretically assess the resulting design and position it in the context of a possible application
in the context of data journalism.

\subsection{Lowering Barriers to Entry}
\label{sec:design-bar}

Data exploration has a certain irreducible essential complexity. To make it accessible to users who
cannot dedicate much time to learning a tool prior to using it, this complexity needs to be
carefully stratified. The Gamma uses a two-level structure. At the first level, the user needs
to learn iterative prompting to be able to start exploring data. At the second level, users will
need to learn domain specific languages defined by individual type providers.

\subsubsection*{Iterative Prompting Principle}
Iterative prompting is a suitable first level principle, because it is easy to trigger.
In conventional programming languages, auto-complete assistance is only available once basic
code structure is written. In contrast, user of In The Gamma only needs to choose the initial
data source. Iterative prompting is also easier to use than e.g.~a command line or a REPL
(read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
The users are not required to recall and type a command. They merely need to select one from a
list of options.

\subsubsection*{Stratifying Data Exploration Complexity}
At the first level, any data source is accessed through the same mechanism. At the second level,
each data source defines its own domain specific language, consisting of the primitives that are
offered to the user in auto-complete prompts. The domain specific languages are embedded in The Gamma
language -- they define merely the available members and cannot extend the core language syntax.
Although the complexity of individual languages differs, the users can always start exploring
and learning new languages using the familiar first level iterative prompting principle.
An important question, which we study in Section~\ref{sec:study}, is whether the expertise gained
with one data source be transferred to working with another data source.

\subsection{Learning without Experts}
\label{sec:design-expert}

A typical user of The Gamma will not dedicate significant time to learning it in advance and
they will not have access to experts. Most learning thus needs to happen from examples.
When analysing how Excel users learn, Sarkar \cite{learning} points out that users learn new
features when the usage of a feature is apparent in a spreadsheet. For example, users can learn
different functions in formulas, because those are visible in the cell. Learning how to use a
wizard for creating charts is not possible because the operation leaves no full trace in the
spreadsheet. Sarkar's recommendation is to \emph{design for percolation},
i.e.~in a way where looking at the final result makes it apparent what feature has been used and how.

In The Gamma, each step of iterative prompting results in an identifier that is
added to the source code. This means that a program constructed solely through iterative prompting
keeps a full trace of how it was created. The resulting source code provides the user all
information that they need to recreate the program, not just by copying it, but also by using the
iterative prompting mechanism. We assess the viability of this way of learning in the study in
Section~\ref{sec:study}.

\subsection{Correctness and Completeness}
\label{sec:design-cc}

An important characteristic of our design is that, barring a few exceptions discussed below, the iterative
prompting mechanism is both correct and complete with respect to possible data exploration scripts.
This means that (i) auto-complete lists offered by iterative prompting contain
only options that are valid in a given context and that (ii) any script that can be written
in The Gamma can be constructed via iterative prompting.

>>>>

\begin{document}
\title{Iterative Prompting: Programmatic Data Exploration for Non-programmers}

@@

\begin{abstract}
  Data exploration tools based on code have many desirable characteristics. They can easily access
  a wide range of different data sources, result in reproducible scripts and encourage users to
  reuse and modify existing code. Unfortunately, most programming tools require expert
  coding skills. Can we make data exploration based on code accessible to non-experts?
  %
  We present The Gamma, a novel text-based data exploration environment that answers the question in
  the affirmative. The Gamma is based on a novel interaction principle, \emph{iterative prompting},
  which lets users create transparent and reproducible scripts without writing code. The Gamma lowers
  the barrier to entry and learning from previously created data analyses. We evaluate the usability
  of The Gamma through a user study on non-technical employees of a research institute.
  %
  Our work shows that we may not need to shy away from code in order to build accessible,
  reproducible and transparent tools that will allow a broad audience to benefit from the
  rise of open data.
\end{abstract}

@@

\begin{figure}
\includegraphics[width=.63\columnwidth]{figures/thegamma-annot}
\vspace{-0.5em}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechanism
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\vspace{-0.5em}
\end{figure}

@@

\section{Introduction}
Despite the advances on visual tooling, programmatic data exploration remains the choice of
expert analysts. It is flexibile, offers greater reusability and leads to transparent
analyses. We aim to make programmatic data exploration accessible to a wider range of users.
The design of a data exploration tool that would make this possible poses a number of challenges.
First, the tool needs to have a low barrier to entry. Second, it needs to support a wide
range of data sources in a uniform way. Third, the users should be able to learn how to use the
tool by looking at existing data analyses.

@@

\subsubsection*{Iterative Prompting}
The main contribution of our work is the \emph{iterative prompting} interaction principle,
which makes it possible to construct all valid data exploration scripts by repeatedly choosing
an item from a list of options offered through auto-complete. The design favors \emph{recognition
over recall} and allows non-programmers to write entire scripts without typing code and without
learning a programming language first. Yet, it still results in transparent and reproducible code.
In other words, iterative prompting turns auto-complete from a programmer
assistance tool into a non-expert programming mechanism. A crucial feature is that iterative
prompting only offers operations that are valid in a given context and that it offers all such
operations; it is both correct and complete.

\subsubsection*{Data Exploration}
The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
The user accesses data available in a structured format. They make several experiments to find an
interesting way of looking at the data, e.g.~by applying different aggregations or filters. They
may choose to view the results as a table or a basic chart before publishing their analysis. The
Gamma makes such programmatic data exploration simple enough for non-experts, but scraping and
cleaning of messy data or building custom data visualizations is currently outside of the scope of
our work. Exposing those using iterative prompting remains an interesting and worthwhile future
challenge.

\subsubsection*{Paper Overview}
The Gamma is available (non-anonymously) at \url{http://thegamma.net},
both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
evaluate the design principles behind the project\footnote{This paper is 11 pages (5500 words),
without citations, to be commensurate with the size of the contribution.}:

@@

\item We evaluate the system through a number of case studies (Section~\ref{sec:cases})
  and a user study (Section~\ref{sec:study}), which confirms  that non-programmers can use The Gamma
  to construct non-trivial data queries.

\item We reflect how our design lowers barriers to entry, supports learning without experts and offers a
  complete and correct program construction method (Section~\ref{sec:discuss}).
\end{itemize}

@@

\noindent
We start from the \ikvd{Doctor} node and then follow two relationships. We use
\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
to follow links to all enemies of the Doctor and then specify
\ikvd{\textquotesingle APPEARED IN\textquotesingle}
to select only enemies that appear in a specific episode. The result appears in
in Figure~\ref{fig:graphtp}.
%
The provider works with any graph database and generates members automatically, based on the
data. In the above, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
relationships. The results returned by the provider is a table of properties of all nodes
along the specified path. As illustrated by an example discussed in Section~\ref{sec:cases},
the returned table can be further queried using the tabular data type provider.

\begin{figure}[t]
\vspace{-0.5em}
\centering
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=0.96\columnwidth]{figures/cases-dr}
  \caption{Exploring Dr Who graph database by composing type providers}
  \label{fig:cases-dr}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=0.96\columnwidth]{figures/cases-mp}
  \caption{Exploring Olympic medallists using tabular data type provider}
  \label{fig:cases-mp}
\end{subfigure}
\vspace{-0.5em}
\caption{Charts produced by two case studies of using The Gamma.}
\label{fig:cases}
\vspace{-0.5em}
\end{figure}

@@

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data explorations. In this section, we consider two case
studies that evaluate expressivity and show what can be achieved using the simple
iterative prompting principle\footnote{Available (non-anonymously) at:
\url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
We used The Gamma for larger projects exploring the UK government expenditure, activities of a
research institute adn Olympic medal winners\footnote{Available (non-anonymously)
at \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.

@@

Using the characterization by Olsen~\cite{evaluating}, data exploration environments are complex
systems that do not yield to simple controlled experimentation. Consequently, our goals are modest and
we do not attempt to quantitatively compare our work with other tools. We merely aim to study whether
The Gamma can be successfully used by non-programmers.

We conducted a study in which we gave volunteers one of four data exploration tasks and assessed
whether they were able to complete the task and how much assistance, if any, they needed.
Some aspects of the study offer insights into how users learn and understand The Gamma. We summarize
those points, but do not claim conclusive results.

\subsection{Study Design}

We performed a between-subjects study to evaluate the first experience of using The Gamma.
We recruited 13 participants (5 male, 8 female) from a business team
of a research institute working in non-technical roles (project management,
partnerships, communications). Only one participant (\#12) had prior programming experience.
We split participants into 4 groups and asked each group to complete a different task.
We gave participants a brief overview of The Gamma (content depended on the task).
The participants worked for 30 minutes, after which we conducted
a 30 minute semi-structured group interview. We let participants work independently, but offered
guidance if they got stuck. The four tasks were:
% The participants were volunteers who responded to an invitation sent to an internal mailing list.

\begin{itemize}
\item \emph{Expenditure.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{expenditure} data source to compare the UK government spending
  on ``Public order and safety'' and ``Defence'' in terms of GDP.
\item \emph{Lords.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{lords} data source (a table with House of Lords
  members expenses) to find a members representing London with the highest travel costs.
\item \emph{Worldbank.} Participants were given a minimal demo of iterative prompting and
  a code sample using \emph{worldbank}. They were asked to solve a different task using
  the \emph{worldbank} data source.
\item \emph{Olympics.} Participants were given a basic demo using \emph{olympics}.
  They were asked to solve a more complex problem, involving grouping and aggregation,
  using the same data source.
\end{itemize}
%
%Some aspects of the study aim to shed light on three questions concerning learnability
%of The Gamma, in particular whether knowledge can be trasferred between different data sources
%(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
%(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
%language used in the tabular data source (\emph{RQ2c}).
%

@@

\noindent
Our primary hypothesis was that non-programmers will be able to use The Gamma to explore
data. This was tested by all four tasks for one of the supported data sources. Some aspects of
the study shed light on questions concerning the learnability of The Gamma. The tasks
\emph{expenditure} and \emph{lords} test if knowledge can be transferred between different
data sources by using one sources in the introduction and another in the task. The task
\emph{worldbank} explores whether users can learn from just code samples by providing only minimal
upfront explanation. The task \emph{lords} lets us study to what extent participants form a
correct mental model of the more complex query language used in the tabular data source.

\begin{table}
\centering
\begin{tabular}{l l l c l}
  \toprule
    & {\small \textit{Task}}
    & {\small \textit{Kind}} & {\small \textit{Done}}
    & {\small \textit{Notes}} \\
  \midrule
  \small \#1\qquad\qquad & \small expenditure \qquad\qquad & \small cube \qquad\qquad & \qquad\priority{50}\qquad\qquad & {\small Obtained one of two data series}\\
  \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series independently}\\
  \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series independently}\\
  \small \#4 & \small expenditure & \small cube & \priority{75}& {\small Completed following a hint to use another member }\\
  \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series independently}\\
  \small \#6 & \small worldbank & \small cube & \priority{75} & {\small Completed after a syntax hint about whitespace}\\
  \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
  \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Completed, but needed longer to find correct data }\\
  \small \#9 & \small lords & \small table & \priority{75} & {\small Struggled with composition of operations}\\
  \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
  \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations taking  arguments}\\
  \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations taking arguments}\\
  \small \#13 & \small olympics & \small table & \priority{75}  & {\small With hints about `then' and operations taking arguments}\\
  \bottomrule
\end{tabular}

@@

\subsection{Study Results}
Table~\ref{tab:tasks} summarizes the work done by the study participants. For each participant, we
record the task, the kind of data source used and the level of completion. For
participants who needed assistance, the notes section details the help given. The experience
suggests a number of possible design improvements for The Gamma, which are discussed below.

\subsubsection*{Can non-programmers explore data with The Gamma?}
All participants were able to complete, at least partially, a non-trivial data exploration task and
only half of them required further guidance. Participants spent 10--25 minutes (average 17 minutes)
working with The Gamma and 12 out of 13 completed the task; 6 required assistance, but 3 of those
faced an issue related to operations taking arguments (discussed later), which could be addressed
in the introduction. A number of participants also shared positive comments in the interviews.
For example, participant \#3 noted taht \emph{``this is actually pretty simple to use.
You think about the logic of what you're actually asking and then you try to get it into the format you can.''}
Participant \#2 noted that The Gamma alleviated their unease about code:
\emph{``For somebody who does not do coding or programming, this does not feel that daunting.
  It's not like you're giving me large screen full of code, which is reassuring.''}

@@

\subsubsection*{How users learn The Gamma?}
There is some evidence that knowledge can be transferred between different data sources. Two of the
tasks (\emph{expenditure} and \emph{lords}) used different data sources in the introduction and the
task. Participants were able to complete those, although \emph{lords} has been challenging as it
involves a complex data source. Participant \#2 also shared a positive comment:
\emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.''}.

There is also some evidence that users can learn just from code samples. In the
\emph{worldbank} task, participants were given only a minimal demo of how to invoke iterative
prompting together with print-out of 2 code samples. All three participants were able to complete
a related task using the same data source. When discussing suitable educational materials for
The Gamma, participant \#7 also confirmed that having code is sufficient when they noted that
\emph{``a video would just be this [i.e.~a code sample] anyway''}. This supports our hypothesis
that, once a user understands the iterative prompting principle, they can learn how to use any
specific data source just from code samples.

\subsubsection*{How users understand complex query languages?}
The tabular type provider has a rich structure and uses a member named \ikvd{then} to complete
the specification of a current operation, for example when specifying a list of aggregation
operations. We asked participants who worked with tabular data (\emph{lords} or \emph{olympics})
about their understanding of the \ikvd{then} member. Two participants (\#12 and \#13) initially
thought that \ikvd{then} is used to split a command over multiple lines, but rejected the idea after
experimenting. Participant \#12 then correctly concluded that it \emph{``allows us
to chain together the operations''} of the query; after a hint, participant \#13 reflected
that \emph{``if I knew this from the start, it would [have been easier].''}
In summary, iterative prompting allows users to start exploring new data sources, but
the structures exposed by more complex data sources have their own further design principles that
the users need to understand.

\subsubsection*{What would make The Gamma easier to use?}
The type provider for tabular data generates operations that take arguments such as \ikvd{take(5)}
in a number of places. When filtering data, it allows writing
\ikvd{olympics.\textquotesingle filter data\textquotesingle.\textquotesingle Year is greater than\textquotesingle(2004)}.
Those operations violate the iterative prompting principle as one cannot type `.' after
\ikvd{\textquotesingle Year is greater than\textquotesingle}. Three participants (\#11, \#12, \#13)
struggled to complete a task, because they initially attempted to use those operations.
This suggests that we should either avoid such operations, or hide them under an ``advanced operations''
tab.

The Gamma uses an ordinary text editor. This has both benefits
and drawbacks compared to structured editors~\cite{structure-based,livenut,lamdu}.
Most participants had no difficulty navigating around code, making edits or deleting
fragments, which is arguably harder in a structured editor. Some participants used
text editor effectively, e.g.~participant \#5, who used copy-and-paste to fetch the same data
for multiple countries. However, we also observed two issues. Participant \#2 struggled
with indentation and participant \#6 had a syntax error in an unrelated command, which
prevents charts from rendering.

@@

\section{Discussion}
\label{sec:discuss}

The Gamma examines an unexplored point in the design space of tools for data exploration. It is
a text-based programming environment for non-programmers. Its design has been motivated
by a curiosity as to whether iterative prompting can make text-based programming with data
accessible to non-experts. In this section, we theoretically assess the resulting design,
making use of criteria for judging whether a system advances the state of the art proposed by
Olsen~\cite{evaluating}.

\subsubsection*{Learning without experts}
The Gamma allows non-programmers to produce transparent and reproducible scripts that explore
data from a wide range of sources. It allows new participants to benefit from the capabilities
offered by programmatic data exploration, satisfying the \emph{exmpowering new participants}
criteria proposed by Olsen~\cite{evaluating}. To empower new participants, our design aims to
make The Gamma suitable for users who cannot dedicate significant amount of time to learning
it in advance and may not have access to experts. This is supported in two ways.

First, the iterative prompting principle makes it easy for users to start experimenting.
The user needs to select an initial data source and then repeatedly choose an
item from a list of choices. Iterative prompting is easier to use than e.g.~a command line or a REPL
(read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
The users are not required to recall and type a command. They merely need to select one from a
list of options. Second, the resulting source code serves as a trace of how the analysis was
created. It provides the user with all information that they need to recreate the
program, not just by copying it, but also by using iterative prompting. Such design
has been termed \emph{design for percolation} by Sarkar \cite{learning}, who studies how Excel users
learn. He points out that users learn new features when the usage of a feature is apparent in a spreadsheet.
For example, users can learn different functions in formulas, because those are visible in the cell.
Learning how to use a wizard for creating charts is hard because the operation leaves no
full trace in the spreadsheet. Design for percolation thus supports learnability.

\subsubsection*{Lowering barriers to entry}

Data exploration has a certain irreducible essential complexity. To make a system usable, this
complexity needs to be carefully stratified. The Gamma uses a two level structure. The first level
consists of the language itself with the iterative prompting mechanism. The second level consists of
the individual members generated by a type provider. This can be seen as a \emph{domain specific
language}, embedded in The Gamma language. Although the complexity of individual domain specific
languages differs, the user can always start exploring through iterative prompting, even when
faced with an unfamiliar data source.
%
In tackling complexity, The Gamma satisfies two criteria proposed by Olsen \cite{evaluating}.
The Gamma satisfies the \emph{generality} criteria in that it can be used uniformly with a wide
range of data sources. It also satisfies the \emph{expressive leverage} criteria in that it factors
out common aspects of different data queries into the core language (first level) and leaves the
specifics of each data source to the second level.

\subsubsection*{Correctness and completeness}

An important characteristic of our design is that the iterative prompting mechanism is both \emph{correct}
and \emph{complete} with respect to possible data exploration scripts. The two properties are
a consequence of the fact that a program is a formed by a chain of operations and
that the auto-completion leverages a static type system. When invoking iterative prompting
at the end of a well-typed script, a selected option, which is a valid object member, is added to
the end of the script, resulting in another well-typed script. This distinguishes our system from
auto-completion based on machine learning, which may offer members not
valid in a given context. Auto-completion lists offered via iterative prompting contain
all available members and so the user can construct all possible scripts. Two exceptions to
completeness in our current design are the let binding and specifying numerical parameters as in
\ikvd{take(5)}.
